---
title: "Project4"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

In this project, you are going to explore matrix factorization methods for recommender system. The goal is to match consumers with most appropriate products. Matrix factorization methods characterize both items and users by vectors of factors inferred from item rating patterns. High correspondence between item and user factors leads to a recommendation. Matrix factorization generally has 3 parts:

- factorization algorithm

- regularization

- postpocessing

It is highly recommended to read this [review paper](./paper/P1 Recommender-Systems.pdf).

### Step 1 Load Data and Train-test Split
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(foreach)
library(doParallel)
library(zoo)
library(Matrix)
library(lubridate)

source("../lib/Matrix_Factorization.R")
source("../lib/cross_validation.R")

data <- read.csv("../data/ml-latest-small/ratings.csv")

set.seed(0)

dataSplit <- train_test_split(data,0.8) #this function ensures that all users and movies are in the training set. 
data_train <- dataSplit$train
data_test <- dataSplit$test
```

###Step 2 Matrix Factorization
#### Step 2.1 Algorithm and Regularization
Here I perform stochastic gradien descent to do matrix factorization.

- For algorithms, the referenced paper are:

A1. [Stochastic Gradient Descent](./paper/P1 Recommender-Systems.pdf) Section: Learning Algorithms-Stochastic Gradient Descent

A2. [Gradient Descent with Probabilistic Assumptions](./paper/P3 probabilistic-matrix-factorization.pdf) Section 2

A3. [Alternating Least Squares](./paper/P4 Large-scale Parallel Collaborative Filtering for the Netflix Prize.pdf) Section 3.1

- For regularizations, the referenced paper are:

R1. [Penalty of Magnitudes](./paper/P1 Recommender-Systems.pdf) Section: a Basic Matrix Factorization Model

R2. [Bias and Intercepts](./paper/P1 Recommender-Systems.pdf) Section: Adding Biases

R3. [Temporal Dynamics](./paper/P5 Collaborative Filtering with Temporal Dynamics.pdf) Section 4 linear model


```{r}
registerDoParallel(cores = detectCores())# use parallelization for faster procesing

out <- als.t(f = 10,  lambda = 0.1,max.iter = 1,data = data, train = data_train, test = data_test)
```


#### Step 2.2 Parameter Tuning
Here you should tune parameters, such as the dimension of factor and the penalty parameter $\lambda$ by cross-validation.
```{r}
source("../lib/cross_validation.R")
f_list <- seq(10, 20, 10)
l_list <- seq(-2, -1, 1)
f_l <- expand.grid(f_list, l_list)
```

```{r, eval=FALSE}
result_summary <- array(NA, dim = c(nrow(f_l), 10, 4)) 
run_time <- system.time(for(i in 1:nrow(f_l)){
    par <- paste("f = ", f_l[i,1], ", lambda = ", 10^f_l[i,2])
    cat(par, "\n")
    current_result <- cv.function(data, K = 5, f = f_l[i,1], lambda = 10^f_l[i,2])
    result_summary[,,i] <- matrix(unlist(current_result), ncol = 10, byrow = T) 
    print(result_summary)
  
})

save(result_summary, file = "../output/rmse.Rdata")
```

```{r}
load("../output/rmse.Rdata")
rmse <- data.frame(rbind(t(result_summary[1,,]), t(result_summary[2,,])), train_test = rep(c("Train", "Test"), each = 4), par = rep(paste("f = ", f_l[,1], ", lambda = ", 10^f_l[,2]), times = 2)) %>% gather("epoch", "RMSE", -train_test, -par)
rmse$epoch <- as.numeric(gsub("X", "", rmse$epoch))
rmse %>% ggplot(aes(x = epoch, y = RMSE, col = train_test)) + geom_point() + facet_grid(~par)
```

### Step 3 Postprocessing
After matrix factorization, postporcessing will be performed to improve accuracy.
The referenced papers are:

P1:[Global bias correction](./paper/P4 Large-scale Parallel Collaborative Filtering for the Netflix Prize.pdf) Section 4.1

P2:[Postprocessing SVD with KNN](./paper/P2 Improving regularized singular value decomposition for collaborative filtering .pdf) Section 3.5

P3:[Postprocessing SVD with kernel ridge regression](./paper/P2 Improving regularized singular value decomposition for collaborative filtering .pdf) Section 3.6

P4:[Linearly combination of predictors](./paper/P4 Large-scale Parallel Collaborative Filtering for the Netflix Prize.pdf) Section 4.1

#### P2 - Postprocessing SVD with KNN
```{r}
source(file = "../lib/p2_svd_knn.R")
## calculate similarity matrix 
q <- out$q
pred_train <- pred_rating_svd_knn(data_train = data_train, data_test = data_train, q)
pred_test <- pred_rating_svd_knn(data_train = data_train, data_test = data_test, q)
train_rmse_p2 <- RMSE2(data_train['rating'], pred_train)
test_rmse_p2 <- RMSE2(data_test['rating'], pred_test)
print(train_rmse_p2)
print(test_rmse_p2)
```

#### P3 - Postprocessing SVD with kernel ridge regression 
```{r}
library("listdtr")
movieID <- colnames(result$q)
MSE <- list()
for(i in 1:610){
  userID = as.character(i)
  print(userID)
  movie_train_index <- which(movieID %in% data_train$movieId[which(data_train$userId == userID)])
  movie_test_index <- which(movieID %in% data_test$movieId[which(data_test$userId == userID)])
  obj <- krr(t(norm_q[, movie_train_index]), data_train$rating[movie_train_index])
  xnew <- norm_q[, movie_test_index]
  ynew <- predict(obj, t(xnew))
  ytrue <- data_test$rating[which(data_test$userId == userID)]
  MSE_ <- mean((ynew - ytrue)^ 2)
  MSE <- append(MSE, MSE_)
}
MSE <- mean(unlist(MSE))
```

### Step 4 Evaluation
You should visualize training and testing RMSE by different dimension of factors and epochs ([One Epoch is when an ENTIRE dataset is passed forward and backward through the neural network only ONCE](https://towardsdatascience.com/epoch-vs-iterations-vs-batch-size-4dfb9c7ce9c9)). 

```{r}
library(ggplot2)

RMSE <- data.frame(epochs = seq(10, 100, 10), Training_MSE = result$train_RMSE, Test_MSE = result$test_RMSE) %>% gather(key = train_or_test, value = RMSE, -epochs)

RMSE %>% ggplot(aes(x = epochs, y = RMSE,col = train_or_test)) + geom_point() + scale_x_discrete(limits = seq(10, 100, 10)) + xlim(c(0, 100))

```
